---
title: "HOILD Demonstration"
author: "Anish Mukherjee and Jeremy T Gaskins"
date: 06/03/2026
fontsize: 10pt
geometry:
  - top=1in
  - bottom=1in
  - left=1in   # Margins can be set in inches (in)
  - right=1in
execute:
  cache.extra: !expr list(file.mtime('HOILD.R'))
format: 
  pdf:
    keep-tex: true
    latex-engine: xelatex
    number-sections: true
    toc: true               # <-- 1. Enables the Table of Contents
    toc-depth: 2            # <-- 2. Optional: Sets how deep the menu nesting goes (1.1.1)
    toc-title: "Contents"
    code-block-border-left: false      # Removes the left vertical bar line
    code-block-bg: "#F7F7F7" 
    include-in-header:
      text: |
        \setlength{\defaultaddspace}{0pt}
        % 1. Force both Code Chunks AND Console Outputs to be footnotesize
        \AddToHook{env/Highlighting/begin}{\small}
        \AddToHook{env/verbatim/begin}{\small}
        
        % 1. Reduce line spacing INSIDE the code and output blocks
        \usepackage{etoolbox}
        \makeatletter
        \preto{\@verbatim}{\baselineskip=10pt\relax} 
        \makeatother
        
        % 2. Shrink vertical gaps BETWEEN chunks and text
        \setlength{\topsep}{0pt}
        \setlength{\partopsep}{0pt}
        \setlength{\parsep}{0pt}
editor: visual
mainfont: "Helvetica"
---

```{r}
#| include: false

knitr::opts_chunk$set(echo = FALSE)
options(knitr.kable.NA = '-')
options(scipen = 999)

library(kableExtra)
library(HDInterval)

n_sim <- 10000
n_burn <- 6000

```

# Introduction

This document demonstrates how the code is used in the manuscript titled "Covariate Informed Identification of Heterogeneity and Outliers in Longitudinal Data" by Anish Mukherjee and Jeremy T Gaskins. The compute-intensive MCMC algorithms and WAIC calculation are implemented in `C++` and have been included in the file `utilities.cpp`, and additional utility functions are included in the file `utilities.R`. To have easy access to these core functions, a set of `R` wrapper functions have been defined in the file `HOILD.R`.

In the following sections, we demonstrate the identification of heterogeneity and outliers in a longitudinal dataset. In @sec-data, we will discuss the format a dataset should have to be used for model fitting. All functions for fitting models HOM-HOV, HEM-HOV, HOM-HOV-O, HEM-HOV-O, and HEM-HEV-O are discussed in @sec-model-fit. The functions for calculating M-WAIC and C-WAIC are presented in @sec-waic. Finally, @sec-estimation evaluates the estimation performance of $\symbf{\beta}$, $\symbf{\gamma}_{u}$, $\symbf{\gamma}_{w}$, and $\symbf{\gamma}_{z}$.

# Data Structure {#sec-data}

We will use CD4 data to demonstrate our approach.

```{r}
#| include: true
#| echo: true
#| cache: true

load('long_dat_all.RData')
ls(log_cd4_dat)

```

The data is a `list` with three elements -- (1) data is a `list`, where the $i$-th element is the `array` of measurements for the $i$-th individual, (2) `time` is a `list` of `array`s where the index $i$ stores the measurement occasions (recoreded in weeks), (3) `fixed.eff` is a `list` of matrices, where the $i$-th element stores the design matrix for the $i$-th individual. The first column in the design matrices represent the intercept.

```{r}
#| include: true
#| echo: true
#| cache: true

dat <- log_cd4_dat
n <- length(dat$data)           # number of individuals
ni_arr <- sapply(dat$data, length)  # an array where ni[i] = number of time-points for i-th individual
ni_max <- max(ni_arr)               # maximum number of measurement occassions
dat$data[1:2]
dat$time[1:2]
dat$fixed.eff[1:2]

```

# Modeling Fitting {#sec-model-fit}

We source the file `HOILD.R` that will load functions `HOM_HOV`, `HEM_HOV`, `HOM_HOV_O`, `HEM_HOV_O`, and `HEV_HEV_O` for fitting the respective models. All these functions generate a single chain of MCMC samples of prespecified lengths. To obtain multiple chains for a model, the corresponding function can be run multiple times (preferably in parallel).

```{r source_code}
#| include: true
#| echo: true

source('HOILD.R')

```

## Model HEM-HEV-O using \texttt{function HEM\_HEV\_O} {#sec-model-fit-hemhevo}

The function takes the following arguments:

-   `dat`: the data formatted as discussed in @sec-data.
-   `fixed_eff_ind`: column indices for the design matrix to be considered as fixed effects
-   `rand_eff_ind`: column indices for the design matrix to be considered as random effects
-   `fixed_eff_u_ind`: column indices for the design matrix to be considered as fixed effects
-   `fixed_eff_w_ind`: column indices for the design matrix to be considered as fixed effects
-   `fixed_eff_z_ind`: column indices for the design matrix to be considered as fixed effects
-   `eta_u`: the scaling factor $\eta_u$ for mean heterogeneity
-   `eta_w`: the scaling factor $\eta_w$ for mean heterogeneity
-   `eta_z`: the scaling factor $\eta_z$ for mean heterogeneity
-   `n_sim`: total number of MCMC iterations
-   `n_burn`: number of iterations to be removed as burnin

The function returns a `list` consisting of MCMC samples for all the parameters. We only keeps samples starting from `n_burn+1` until `n_sim`, resulting in a sample size of `(n_sim-n_burn+1)`. The parameters returned are:

-   `beta`: a matrix of posterior samples of $\symbf{\beta}$. It has size `(n_sim-n_burn+1) x p`, where `p = length(fixed_eff_ind)` is the number of fixed effects including the intercept.
-   `r_mat`: a matrix of posterior samples of $\symbf{r}_i$, where $i = 1, \ldots, n$. It has size `(n_sim-n_burn+1) x (nq)`, where `q = length(rand_eff_ind)` is the number of random effects per individual and `n` is the total number of individuals.
-   `sigma2`: a matrix of posterior samples of $\symbf{\sigma}^2 = (\sigma_0^2, \sigma_1^2)$. It has size `(n_sim-n_burn+1) x 2`. The first column has $\sigma_0^2$ samples, and the second has $\sigma_1^2$ samples.
-   `sigma.e2`: a matrix of posterior samples of $\sigma_i^2$, where $i = 1, \ldots, n$. It has size `(n_sim-n_burn+1) x n`.
-   `alpha`: a matrix of posterior samples. It has size `(n_sim-n_burn+1) x 2`, where the first columns consists if 0, and the second column represents the $\alpha$ for the heterogeneous group.
-   `rho`: an array of posterior samples of $\rho$. It has length `(n_sim-n_burn+1)`.
-   `u`: a matrix if posterior samples of $\symbf{u} = (u_1, \ldots, u_n)$, the mean heterogeneity indicators. It has size `(n_sim-n_burn+1) x n`.
-   `w`: a matrix if posterior samples of $\symbf{w} = (\symbf{w}_1, \ldots, \symbf{w}_n)$, where $\symbf{w}_i = (w_{i1}, \ldots, w_{in_i})$ is the indicators for the observation-level outliers for $i$-th individual. It has size `(n_sim-n_burn+1) x (n*ni_max)`.
-   `z`: a matrix if posterior samples of $\symbf{z} = (z_1, \ldots, z_n)$, the variance heterogeneity indicators. It has size `(n_sim-n_burn+1) x n`.
-   `gamma.u`: a matrix of posterior samples of $\symbf{\gamma}_u$. It has size `(n_sim-n_burn+1) x d_u`, where `d_u = length(fixed_eff_u_ind)` is the number of predictors in the logistic model for the mean heterogeneity indicators.
-   `gamma.w`: a matrix of posterior samples of $\symbf{\gamma}_w$. It has size `(n_sim-n_burn+1) x d_w`, where `d_w = length(fixed_eff_w_ind)` is the number of predictors in the logistic model for the observation-level outliers.
-   `gamma.z`: a matrix of posterior samples of $\symbf{\gamma}_z$. It has size `(n_sim-n_burn+1) x d_z`, where `d_z = length(fixed_eff_zind)` is the number of predictors in the logistic model for the variance heterogeneity indicators.

We use the `function HEM_HEV_O` to identify heterogeneity and outliers in the CD4 data as follows,

```{r hem_hev_o}
#| include: true
#| echo: true

mod <- HEM_HEV_O (
  dat,
  fixed_eff_ind = 1:8,
  rand_eff_ind = 1:2,
  fixed_eff_u_ind = c(1,4:8),
  fixed_eff_w_ind = c(1,2,4:8),
  fixed_eff_z_ind = c(1,4:8),
  eta_u = 3,
  eta_w = 3,
  eta_z = 3,
  n_sim = n_sim,
  n_burn = n_burn
)
ls(mod)

## saving the model
save(mod, file = "mod_HEM_HEV_O.RData")
```

## Submodels HOM-HOV, HEM-HOV, HOM-HOV-O, HEM-HOV-O

For each of the sub-models a separate model fitting function is defined with a similar set of arguments excluding those which are not relevant for the respective models. All functions return the same `list` of parameters as returned by `HEM_HEV_O`.

\textbf{\texttt{HOM\_HOV}}: The arguments are `dat`, `fixed_eff_ind`, `rand_eff_ind`, `n_sim`, `n_burn`. Among the `list` of parameters returned,

-   `u`, `w`, and `z` will be all zeros since the model HOM-HOV asssumes no heterogeneity
-   `gamma.u`, `gamma.w`, and `gamma.z` samples are drawn from their respective priors and can be ignored.
-   The first column of the `sigma2` are posterior samples of $\sigma_0^2$, while the second column of `sigma2` representing $\sigma_1^2$ are drawn from its prior and can be ignored.
-   The variables `alpha` and `sigma.e2` can be ignored as they are not relevant to the model HOM-HOV.

```{r hom_hov}
#| include: true
#| echo: true

mod <- HOM_HOV (
  dat,
  fixed_eff_ind = 1:8,
  rand_eff_ind = 1:2,
  n_sim = n_sim,
  n_burn = n_burn
)
save(mod, file = "mod_HOM_HOV.RData")
```

\textbf{\texttt{HEM\_HOV}}: The arguments are `dat`, `fixed_eff_ind`, `rand_eff_ind`, `fixed_eff_u_ind`, `eta_u`, `n_sim`, `n_burn`. Among the parameters returned by the function,

-   `u` will be sampled from its posterior distribution, while `w` and `z` will be all zeros indicating only mean heterogeneity
-   `gamma.u` will be posterior samples of $\symbf{\gamma}_u$, and we can ignore `gamma.w` and `gamma.z` variables.
-   The interpretation of `sigma2` is same as in `HOM_HOV`.
-   The variables `alpha` and `sigma.e2` can be ignored as they are not relevant to the model HEM-HOV.

```{r hem_hov}
#| include: true
#| echo: true

mod <- HEM_HOV (
  dat,
  fixed_eff_ind = 1:8,
  rand_eff_ind = 1:2,
  fixed_eff_u_ind = c(1,4:8),
  eta_u = 3,
  n_sim = n_sim,
  n_burn = n_burn
)
save(mod, file = "mod_HEM_HOV.RData")
```

`HOM_HOV_O`: The arguments are `dat`, `fixed_eff_ind`, `rand_eff_ind`, `fixed_eff_w_ind`, `eta_w`, `n_sim`, `n_burn`. Among the parameters returned by the function,

-   `w` will be sampled from its posterior distribution, while `u` and `z` will be all zeros allowing only observation-level outliers
-   `gamma.w` will be posterior samples of $\symbf{\gamma}_w$, while `gamma.u` and `gamma.z` can be ignored.
-   The interpretation of `sigma2` is same as in `HOM_HOV`.
-   The variables `alpha` and `sigma.e2` can be ignored as they are not relevant to the model HOM-HOV-O.

```{r hom_hov_o}
#| include: true
#| echo: true

mod <- HOM_HOV_O (
  log_cd4_dat,
  fixed_eff_ind = 1:8,
  rand_eff_ind = 1:2,
  fixed_eff_w_ind = c(1,2,4:8),
  eta_w = 3,
  n_sim = n_sim,
  n_burn = n_burn
)
save(mod, file = "mod_HOM_HOV_O.RData")
```

`HEM_HOV_O`: The arguments are `dat`, `fixed_eff_ind`, `rand_eff_ind`, `fixed_eff_u_ind`, `fixed_eff_w_ind`, `eta_u`, `eta_w`, `n_sim`, `n_burn`. Among the parameters returned by the function,

-   `u` and `w` will be sampled from their posterior distributions, while `z` will be all zeros due to the assumption of variance heterogeneity
-   `gamma.u` and `gamma.w` will be posterior samples of $\symbf{\gamma}_u$ and $\symbf{\gamma}_w$ respectively, and we can ignore `gamma.z`.
-   The interpretation of `sigma2` is same as in `HOM_HOV`.
-   The variables `alpha` and `sigma.e2` can be ignored as they are not relevant to the model HEM-HOV-O.

```{r hem_hov_o}
#| include: true
#| echo: true

mod <- HEM_HOV_O (
  dat,
  fixed_eff_ind = 1:8,
  rand_eff_ind = 1:2,
  fixed_eff_u_ind = c(1,4:8),
  fixed_eff_w_ind = c(1,2,4:8),
  eta_u = 3,
  eta_w = 3,
  n_sim = n_sim,
  n_burn = n_burn
)
save(mod, file = "mod_HEM_HOV_O.RData")

```

Note that, the `list` contains several parameters, e.g., the indicators `u`, `w`, `z`, `gamma.u`, `gamma.w`, `gamma.z`, `alpha` that are not directly relevant for HOM-HOV model. We made this choice to keep the format of the model objects same across HEM-HEV-O and all its submodels. To extract only the relevant set of parameters for a particular model `extract_params()` `function` is provided that can be used as follows:

```{r extract_parameters}
#| include: true
#| echo: true

## extracting the only the relevant parameters
params_list <- extract_params(mod, "HEM-HEV-O")
ls(params_list)


```

# Model Selection {#sec-waic}

We calculate WAIC for the model HEM-HEV-O using `function compute_WAIC_HEM_HEV_O`. The arguments of the function are:

-   `dat`: the dataset
-   `mod`: the list of parameters obtained by running `function HEM_HEV_O` on `dat`
-   `iter_sel`: the array of iteration indices for the posterior samples to be used to calculate WAIC
-   `type`: can take two values "C-WAIC" and "M-WAIC" representing conditional and marginal version of WAIC
-   rest of the parameters are as discussed in @sec-model-fit-hemhevo

The function returns C-WAIC or M-WAIC depending on the argument `type`.

We use the function to calculate the WAIC for model HEM-HEV-O on CD4 data as follows,

```{r waic}
#| include: true
#| echo: true

load('mod_HEM_HEV_O.RData')
iter_sel <- seq(1,n_sim-n_burn,4)

WAIC <- array(NA, 2)
waic_types <- c("M-WAIC", "C-WAIC")
for (i in seq_along(waic_types)) {
  WAIC[i] <- compute_WAIC_HEM_HEV_O(
    dat,
    mod,
    iter_sel,
    fixed_eff_ind = 1:8,
    rand_eff_ind = 1:2,
    fixed_eff_u_ind = c(1,4:8),
    fixed_eff_w_ind = c(1,2,4:8),
    fixed_eff_z_ind = c(1,4:8),
    eta_u = 3,
    eta_w = 3,
    eta_z = 3,
    type = waic_types[i]
  )
}
save(WAIC, file = "WAIC_HEM_HEV_O.RData")

```
The calculation yields M-WAIC = `r round(WAIC[1],2)`, and C-WAIC = `r round(WAIC[2],2)`.

For the rest of the models, we use the corresponding WAIC `function`s as follows,

```{r waic_hom_hov}
#| include: true
#| echo: true

load('mod_HOM_HOV.RData')
iter_sel <- seq(1,n_sim-n_burn,4)

WAIC <- array(NA, 2)
waic_types <- c("M-WAIC", "C-WAIC")
for (i in seq_along(waic_types)) {
  WAIC[i] <- compute_WAIC_HOM_HOV (
    dat,
    mod,
    iter_sel,
    fixed_eff_ind = 1:8,
    rand_eff_ind = 1:2,
    type = waic_types[i]
  )
}
save(WAIC, file = "WAIC_HOM_HOV.RData")
```
For model HOM-HOV, M-WAIC = `r round(WAIC[1],2)`, and C-WAIC = `r round(WAIC[2],2)`.


```{r waic_hem_hov}
#| include: true
#| echo: true

load('mod_HEM_HOV.RData')
iter_sel <- seq(1,n_sim-n_burn,4)

WAIC <- array(NA, 2)
waic_types <- c("M-WAIC", "C-WAIC")
for (i in seq_along(waic_types)) {
  WAIC[i] <- compute_WAIC_HEM_HOV (
    dat,
    mod,
    iter_sel,
    fixed_eff_ind = 1:8,
    rand_eff_ind = 1:2,
    fixed_eff_u_ind = c(1,4:8),
    eta_u = 3,
    type = waic_types[i]
  )
}
save(WAIC, file = "WAIC_HEM_HOV.RData")
```
For model HEM-HOV, M-WAIC = `r round(WAIC[1],2)`, and C-WAIC = `r round(WAIC[2],2)`.


```{r waic_hom_hov_o}
#| include: true
#| echo: true

load('mod_HOM_HOV_O.RData')
iter_sel <- seq(1,n_sim-n_burn,4)

WAIC <- array(NA, 2)
waic_types <- c("M-WAIC", "C-WAIC")
for (i in seq_along(waic_types)) {
  WAIC[i] <- compute_WAIC_HOM_HOV_O (
    dat,
    mod,
    iter_sel,
    fixed_eff_ind = 1:8,
    rand_eff_ind = 1:2,
    fixed_eff_w_ind = c(1,2,4:8),
    eta_w = 3,
    type = waic_types[i]
  )
}
save(WAIC, file = "WAIC_HOM_HOV_O.RData")
```
For model HOM-HOV_O, M-WAIC = `r round(WAIC[1],2)`, and C-WAIC = `r round(WAIC[2],2)`.

```{r waic_hem_hov_o}
#| include: true
#| echo: true

load('mod_HEM_HOV_O.RData')
iter_sel <- seq(1,n_sim-n_burn,4)

WAIC <- array(NA, 2)
waic_types <- c("M-WAIC", "C-WAIC")
for (i in seq_along(waic_types)) {
  WAIC[i] <- compute_WAIC_HEM_HOV_O (
    dat,
    mod,
    iter_sel,
    fixed_eff_ind = 1:8,
    rand_eff_ind = 1:2,
    fixed_eff_u_ind = c(1,4:8),
    fixed_eff_w_ind = c(1,2,4:8),
    eta_u = 3,
    eta_w = 3,
    type = waic_types[i]
  )
}
save(WAIC, file = "WAIC_HEM_HOV_O.RData")
```

For model HEM-HOV_O, M-WAIC = `r round(WAIC[1],2)`, and C-WAIC = `r round(WAIC[2],2)`.


# Generating Tables and Plots {#sec-estimation}

We now demonstrate how to generate Tables and plots presented in the manuscript. We start by generating a part of Table D.2 in the supplimentary document. The following code generates results for $\eta = 3$, and the code can be suitably adjusted to generate the rest of the information in Table D.2.

```{r heterogeneity_estimation}
#| include: true
#| echo: true

mod_type_arr <- c("HOM_HOV", "HEM_HOV", "HOM_HOV_O", "HEM_HOV_O", "HEM_HEV_O")
indc_mat <- matrix(NA, nrow = 5, ncol = 3)
sigma02_mat <- matrix(NA, nrow = 5, ncol = 3)
sigma12_arr <- array(NA, 5)
alpha_arr <- array(NA, 5)
iter_sel <- seq(1,n_sim-n_burn,4)
for (i in seq_along(mod_type_arr)) {
  load(paste0('mod_', mod_type_arr[i], ".RData"))
  indc_mat[i,] <- c(
    mean(colMeans(mod$u[iter_sel,])),
    mean(colMeans(mod$w[iter_sel,])),
    mean(colMeans(mod$z[iter_sel,]))
  )
  sigma02_mat[i,] <- formatC(
    c(
      mean(mod$sigma2[iter_sel,1]),
      as.numeric(hdi(mod$sigma2[iter_sel,1]))
    ), 
    digits = 3, format = "f"
  )
  if (i == 5) {
    sigma12_arr[i] <- formatC(mean(mod$sigma2[iter_sel,2]), digits=3, format="f")
    alpha_arr[i] <- formatC(mean(mod$alpha[iter_sel,2]), digits=3, format="f")
  }
}
est_df <- cbind.data.frame(
  formatC(round(indc_mat, 3), digits=3, format = "f"), 
  sigma02_mat[,1],
  paste0("(", sigma02_mat[,2], ", ", sigma02_mat[,3], ")"),
  sigma12_arr,
  alpha_arr
)
rownames(est_df) <- gsub("_", "-", mod_type_arr)
colnames(est_df) <- c(
  "$\\widehat{u}$", "$\\widehat{w}$", "$\\widehat{z}$", "$\\widehat{\\sigma_0^2}$", 
  "95\\% CI", "$\\widehat{\\sigma_1^2}$", "$\\widehat{\\alpha}$"
)

kbl(est_df, booktabs=T, align = c("l", rep("c", ncol(est_df)-1)), escape=F) %>%
  kable_styling(position = "center", latex_options=c("hold_position"))

```


Now we generate the profile plots with outliers and mean/variance heterogeneity.

## Generating plot showing observation-level outliers

```{r plots}
#| include: true
#| echo: true
#| fig-width: 5
#| fig-height: 4.5

load('mod_HEM_HEV_O.RData')
fixed_eff_ind <- 1:8
beta_est <- colMeans(mod$beta[iter_sel,])
r_est <- matrix(colMeans(mod$r_mat[iter_sel,]), ncol=2)
w_est <- colMeans(mod$w[iter_sel,])
out_id <- as.numeric(w_est > 0.5)

id_arr_list <- list()
max_w_est_profile <- array(NA, length(dat$data))
out_id_list <- list()
for (i in 1:n) {
  id_arr_list[[i]] <- ni_max*(i-1) + (1:ni_arr[i])
  max_w_est_profile[i] <- max(w_est[id_arr_list[[i]]])
  out_id_list[[i]] <- out_id[id_arr_list[[i]]]
}

n_indv <- 8
rem1 <- order(max_w_est_profile, decreasing=TRUE)[1:n_indv]
rem2 <- rev(order(max_w_est_profile, decreasing=FALSE)[1:n_indv])
rem <- c(rem1, rem2)
out_indc <- do.call(c, out_id_list[rem])

plt_df <- data.frame(
  tt = unlist(lapply(rem, function(i) {dat$time[[i]]})),
  y = unlist(lapply(
    rem, 
    function(i) {c(dat$data[[i]] - dat$fixed.eff[[i]][,fixed_eff_ind] %*% beta_est)}
  )),
  id = factor(
    unlist(lapply(
      rem, function(i) {
        rep(paste0("ID: ", i, " (prob: ", max_w_est_profile[i], ")"), each = ni_arr[i])
      }
    )),
    levels = unlist(lapply(rem, function(i) {
      paste0("ID: ", i, " (prob: ", max_w_est_profile[i], ")")
    }))),
  out = factor(out_indc, levels = c(0,1), labels = c("No", "Yes"))
)

ggplot(data = plt_df, aes (x = tt, y = y)) +
  geom_point(aes(size = out)) +
  scale_size_manual(breaks = c("No", "Yes"), values = c(1,3)) +
  scale_y_continuous(expand = c(0.1, 0.1)) +
  geom_line(color = "darkgrey") +
  labs(x = "time", y = "residuals", size = "Outlier") +
  facet_wrap(~id, nrow = 4, scales="free_x") +
  theme_bw() +
  theme(legend.position = "bottom", strip.background = element_rect(fill = "white"))

```

## Generating plot showing mean heterogeneity

```{r mean_het_plot}
#| include: true
#| echo: true
#| fig-width: 5
#| fig-height: 3

n_indv <- 4
rand_eff_ind <- 1:2
beta_est <- colMeans(mod$beta[iter_sel,])
r_est <- matrix(colMeans(mod$r_mat[iter_sel,]), ncol=2)
u_est <- colMeans(mod$u[iter_sel,])
het_ind <- order(rank(u_est, ties.method = "first"), decreasing = TRUE)[1:(n_indv)]
hom_ind <- order(rank(u_est, ties.method = "first"), decreasing = FALSE)[1:(n_indv)]
u_ind <- c(het_ind, rev(hom_ind))
ni_sum <- c(0, cumsum(sapply(dat$data, function(x) {length(x)})))
# w_prob_list <- lapply(u_ind, function(ind) {w_est[(ni_sum[ind]+1):ni_sum[ind+1]]})
w_prob_list <- lapply(u_ind, function(ind) {w_est[id_arr_list[[ind]]]})
w_prob <- factor (
  ifelse (unlist(w_prob_list) > 0.5, 1, 0),
  levels = c(0,1), labels = c("No", "Yes")
)
max_w_prob <- sapply(w_prob_list, function(l) {max(l)})

ni_ind <- sapply(u_ind, function(i) {length(dat$data[[i]])})

plt_u_dat <- data.frame (
  tt = unlist(lapply(u_ind, function(i) {dat$time[[i]]})),
  y = unlist(lapply(
    u_ind, 
    function(i) {
      c(dat$data[[i]] - dat$fixed.eff[[i]][,fixed_eff_ind] %*% beta_est)
    }
  )),
  id = factor(unlist(lapply(
    1:length(ni_ind),
    function(i) {
      rep(paste0("ID: ", u_ind[i],
                 " (prob: ", round(u_est[u_ind[i]],3), ")"),
          each = ni_ind[i])
    }
  )),
  levels = unlist(lapply(
    1:length(ni_ind),
    function(i) {
      paste0("ID: ", u_ind[i], " (prob: ", round(u_est[u_ind[i]],3) ,")")
    }
  ))),
  r_i = do.call(c, lapply(
    u_ind, function(i) c(dat$fixed.eff[[i]][,rand_eff_ind]%*%r_est[i,])
  )),
  w_prob = w_prob
)

ggplot(
  data = plt_u_dat, aes (x = tt, y = y, r_i = r_i)
) +
  geom_point(aes(size=w_prob)) +
  geom_line(color = "darkgrey") +
  geom_line(aes(x=tt, y=r_i), color="blue") +
  geom_hline(yintercept = 0, color = "red") +
  labs(x = "time", y = "residuals", title = "Subject-level Mean Heterogeneity") +
  scale_y_continuous(expand = c(0.1, 0.1)) +
  scale_size_manual(
    name = "Outlier", breaks = c("No", "Yes"), values = c(1,3), drop = FALSE
  ) +
  facet_wrap(~id, nrow = 2, ncol = 4) +
  theme_bw() +
  theme(legend.position = "bottom",
        strip.background = element_rect(fill = "white"),
        legend.key.width = unit(1, 'cm'),
        plot.title = element_text(size=11))

```

## Generating plot showing variance heterogeneity

```{r var_het_plot}
#| include: true
#| echo: true
#| fig-width: 5
#| fig-height: 3

n_indv <- 4
beta_est <- colMeans(mod$beta[iter_sel,])
r_est <- matrix(colMeans(mod$r_mat[iter_sel,]), ncol=2)
z_est <- colMeans(mod$z[iter_sel,])
het_ind <- order(rank(z_est, ties.method = "first"), decreasing = TRUE)[1:(n_indv)]
hom_ind <- order(rank(z_est, ties.method = "first"), decreasing = FALSE)[1:(n_indv)]
z_ind <- c(het_ind, rev(hom_ind))
ni_sum <- c(0, cumsum(sapply(dat$data, function(x) {length(x)})))
# w_prob_list <- lapply(z_ind, function(ind) {W_est[(ni_sum[ind]+1):ni_sum[ind+1]]})
w_prob_list <- lapply(z_ind, function(ind) {w_est[id_arr_list[[ind]]]})
w_prob <- factor (
  ifelse (unlist(w_prob_list) > 0.5, 1, 0),
  levels = c(0,1), labels = c("No", "Yes")
)
max_w_prob <- sapply(w_prob_list, function(l) {max(l)})

ni_ind <- sapply(z_ind, function(i) {length(dat$data[[i]])})

plt_z_dat <- data.frame (
  tt = unlist(lapply(z_ind, function(i) {dat$time[[i]]})),
  y = unlist(lapply(
    z_ind, 
    function(i) {
      c(dat$data[[i]] - dat$fixed.eff[[i]][,fixed_eff_ind] %*% beta_est)
    }
  )),
  id = factor(unlist(lapply(
    1:length(ni_ind),
    function(i) {
      rep(paste0("ID: ", z_ind[i],
                 " (prob: ", round(z_est[z_ind[i]],3), ")"),
          each = ni_ind[i])
    }
  )),
  levels = unlist(lapply(
    1:length(ni_ind),
    function(i) {
      paste0("ID: ", z_ind[i], " (prob: ", round(z_est[z_ind[i]],3) ,")")
    }
  ))),
  r_i = do.call(c, lapply(
    z_ind, function(i) c(dat$fixed.eff[[i]][,rand_eff_ind]%*%r_est[i,])
  )),
  w_prob = w_prob
)

ggplot(
  data = plt_z_dat, aes (x = tt, y = y, r_i = r_i)
) +
  geom_point(aes(size=w_prob)) +
  geom_line(color = "darkgrey") +
  geom_line(aes(x=tt, y=r_i), color="blue") +
  geom_hline(yintercept = 0, color = "red") +
  labs(x = "time", y = "residuals", title = "Subject-level Variance Heterogeneity") +
  scale_y_continuous(expand = c(0.1, 0.1)) +
  scale_size_manual(
    name = "Outlier", breaks = c("No", "Yes"), values = c(1,3), drop = FALSE
  ) +
  facet_wrap(~id, nrow = 2, ncol = 4) +
  theme_bw() +
  theme(legend.position = "bottom",
        strip.background = element_rect(fill = "white"),
        legend.key.width = unit(1, 'cm'),
        plot.title = element_text(size=11))

```

